Collaborative Visual Analysis on the Web with RCloud

### 2015-03-31 relatedwork.tex
copy editing

\section{Related Work}

In some sense or another, the entire field of information
visualization and visual analytics revolve around improving how users
solve problems using data. In this section, we focus specifically on
system work relating to problem-solving environments and multiple users.

@@

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it here can be traced all the
way back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
literate programming, a comprehensive description in prose of the
behavior of the program is ``weaved'' together with the source code,
yielding both an executable program and a human-readable document.  A
notebook represented as a collection of short, executable cells,
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally implemented as part of
IPython) offer some notebook features, but lack a transparent
mechanism for sharing and deployment in multiple-user settings.
%
Further afield, {\bf Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.
%
Although RStudio offers publication of literate R programs as a free
service on their website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for other users to build off of the work published (or even for the
original author to update new versions). In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.

\paragraph*{Provenance and versioning} As
we mentioned, one central problem in exploratory analysis is that
problems change quickly over time, often in the course of developing a
solution. As a result, systems should provide adequate support for
tracking \emph{changes} of the data analyses scripts.  VisTrails was
one of the original systems for managing \emph{process provenance} ,
and demonstrated the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

@@

\paragraph*{Web-based tools for sharing code snippets}
There have been a variety of tools recently developed for quickly
sharing small programs on the Web, including but not limited to:
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}.  bl.ocks and jsfiddle are both designed to
share Javascript programs, which means that their deployment happens
automatically through the web browser. If and when Javascript becomes
the lingua franca of exploratory data analysis, then we can foresee
building a much simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not a realistic assumption for the present, making these solutions by
themselves unsuitable to our use case. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: in other
words, it is possible to generate a plot.ly visualization from inside
a separate program. Although this is an intriguing idea, it
nevertheless creates a disconnect between the analysis and the
resulting visualization. In RCloud, we wanted to make sure that every
visualization was transparently linked to the source code that
generated it.

@@

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF}.
RCloud notebooks, and the integrated version control system for them,
described in Section~\ref{sec:notebooks}, address modularity and granularity,
and artifact histories.
\emph{Starring}, the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses the cost of
integration, content export, presentation and view sharing.

@@

There has been noteworthy work on specific techniques such as
social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
and crowdsourcing \cite{Fast:2014:ECS} to support collaborative
or social development or analysis processes.
Similarly, there are computational methods to support high
performance execution in incremental code development
environments \cite{Guo:2010:TPI}.
The goal of RCloud is to define an environment in which many such
techniques may be integrated and made available to a broad community.
\begin{figure*}
\centering
\includegraphics[width=.95\linewidth]{fig/notebook/notebook.pdf}
\caption{\label{fig:notebook}An RCloud notebook is a sequence of
  \emph{cells}, each a snippet of source in one of the supported languages (typically R, but Python and others are also supported) or Markdown. The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, providing us with easy access primitives for version tracking. Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL. This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{sec:interviews}. }
\end{figure*}

@@

One overall goal is to reduce the gap between implementers and
deployers in visual analytics. The fusion of development with
production operations in software release management (``DevOps''
\cite{Httermann:2012:DD} or ``continuous integration''
\cite{Fowler:2006:Continuous}) is a trend in web services and related
fields.  By making it convenient for data scientists to expend just a
little more effort when creating experiments, we may be able to
eliminate the need for programming teams to recreate their work to
deploy it in production, which has a high cost in time, expense and
accuracy. We now turn to a description of the system architecture
itself, and how it enables the features we have been highlighting.

>>>>

\section{Related Work\label{sec:related}}

Broadly, the entire field of information visualization and visual analytics
revolves around improving how users solve problems involving data.
In this section, we focus specifically on
system work relating to problem-solving environments and multiple users.

@@

\paragraph*{Notebooks as a medium for data analysis dissemination}
The concept of a ``notebook'' as we use it here can be traced all the
way back to Knuth's literate programming~\cite{Knuth:1984:LP}. In
literate programming, a comprehensive description in prose of the
behavior of the program is ``weaved'' together with the source code,
yielding both an executable program and a human-readable document.
A notebook represented as a collection of short, executable cells,
originated with Mathematica, and in R, literate programming is
supported by packages such as knitr and RMarkdown~\cite{Xie:2013:DDW}.
Project Jupyter~\cite{jupyter} (originally implemented as part of
IPython) offer some notebook features, but lack a transparent
mechanism for sharing and deployment in multiple-user settings.
%
Further afield, {Electronic Lab Notebooks} are applications for organizing
and sharing data from scientific lab experiments\cite{Rubacha:2011:ELN}.
In a sense, we hope to adapt and extend this concept to the work of
visual analytics teams.
%
Although RStudio offers publication of literate R programs as a free
service on their website, the workflow is somewhat disconnected from
the development of those programs. Once they're uploaded, it's hard
for other users to build off of the work published (or even for the
original author to update new versions). In other words, RStudio
handles \emph{publication}, but not \emph{deployment} and sharing.

\paragraph*{Provenance and versioning} As
mentioned, one central issue in exploratory analysis is that
problems change quickly over time, often in the course of developing a
solution. As a result, systems should provide adequate support for
tracking \emph{changes} of the data analysis scripts. VisTrails was
one of the original systems for managing \emph{process provenance} ,
and demonstrated the value of capturing aspects of the processes that
surround data analysis experiments and tools, including detailed
history, collaboration, and deployment~\cite{Callahan:2006:VVM}.

@@

\paragraph*{Web-based tools for sharing code snippets}
There have been quite a few tools recently developed for quickly
sharing small programs on the web, including
bl.ocks~\cite{blocks}, jsfiddle~\cite{jsfiddle}, and
plot.ly~\cite{plotly}. bl.ocks and jsfiddle are designed to
share Javascript programs, which means that deployment happens
automatically through the web browser. If Javascript eventually becomes
the lingua franca of exploratory data analysis, we can foresee
building a simpler version of RCloud where all execution happens
either on the client side or via web services. Unfortunately that is
not a realistic assumption at present, making these stand-alone
solutions unsuitable for our purpose. Plot.ly is notable in that it
provides API support for publishing \emph{from} scripts: in other
words, it is possible to generate a plot.ly visualization from inside
another program. Although this is an intriguing idea, it
nevertheless creates a disconnection between the analysis and the
resulting visualization. In RCloud, we wanted to ensure that every
visualization is transparently linked to the source code that
generates it.

@@

Heer and Agarwala identify many design considerations for
collaborative visual analytics~\cite{Heer:2008:DCF} that
influenced our work.
RCloud notebooks, and the integrated version control system for them,
described in Section~\ref{sec:notebooks}, address modularity and granularity,
and artifact histories.
\emph{Starring}, the means for signaling interest in notebooks, described in
Section~\ref{sec:starring}, addresses social-psychological incentives,
recommendation, and voting and ranking. RCloud's integrated deployment
mechanism, described in Section~\ref{sec:deployment}, addresses the cost of
integration, content export, presentation and view sharing.

@@

There has been noteworthy work on specific techniques
to support collaborative or social code development and data analysis,
such as social bookmarking \cite{Millen:2006:DSB} \cite{Heer:2007:VAV}
and crowdsourcing \cite{Fast:2014:ECS}.
Similarly, there are computational methods to support high
performance execution in incremental code development
environments \cite{Guo:2010:TPI}.
The goal of RCloud is to define an environment in which many such
techniques may be integrated and made available to a broad community.
\begin{figure*}
\centering
\includegraphics[width=.95\linewidth]{fig/notebook/notebook.pdf}
\caption{\label{fig:notebook}An RCloud notebook is a sequence of
\emph{cells}, each a snippet of source in one of the supported languages (typically R, but Python and others are also supported) or Markdown. The main creation workflow involves editing notebooks, which are transparently stored as git repositories in GitHub, providing us with easy access primitives for version tracking. Notebooks can be executed as they're edited (left), or in a standalone viewer (right), via a slightly different URL. This lightweight provides a very low-friction mechanism for sharing results which we discuss in full in Section~\ref{sec:interviews}. }
\end{figure*}

@@

One overall goal is to reduce the gap between implementers and
deployers in visual analytics. The fusion of development with
production operations in software release management (``DevOps''
\cite{Httermann:2012:DD} or ``continuous integration''
\cite{Fowler:2006:Continuous}) is a trend in web services and similar
fields. By making it convenient for data scientists to expend just a
little additional effort when creating experiments, we may be able to
eliminate the need for programming teams to recreate their work to
deploy it in production, which has a high cost in time, expense and
accuracy.

We next give a description of the system architecture,
and how it enables capabilities that satisfy the requirements as described.
